---
title: "04 — Best vs Standard Pipeline Comparison"
subtitle: "Edit Distance Analysis & Integrated Dendrogram"
author: "Andrés Catalán Tatay"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    fig_width: 14
    fig_height: 10
---

<!--
================================================================================
  Project : fliC Amplicon Sequencing for High-Resolution Profiling of E. coli
            in the Infant Gut Microbiome
  Author  : Andrés Catalán Tatay
  Part    : Analysis — Step 04
  Purpose : Direct comparison of the BEST pipeline (Strict filter, Binned
            error model, FALSE pooling) against the STANDARD pipeline (Loose
            filter, PacBio error model, FALSE pooling).
            Covers: edit distance analysis, BLAST-based trust tier assignment,
            integrated dendrogram, and biological group composition by ASV origin.
  Inputs  : data/cases/                  — 96 RDS files (DADA2 outputs)
             data/wgs/ps_wgs_EcfliC.RDS  — WGS phyloseq reference
             data/blast/JSON/             — BLAST Single-file JSON results
                                           (one per FASTA chunk, submitted
                                            manually to NCBI BLAST)
  Outputs : results/04_best_vs_standard/
              04_all_asvs.fasta                     — full FASTA for BLAST
              blast/blast_part_NN.fasta              — split FASTA parts
              blast/JSON/                            — BLAST JSON files (input)
              blast/blast_df_parsed.csv              — parsed BLAST cache
              cached_edit_distance_matrix.rds        — edit distance cache
              Figure1_Edit_Distance_Distribution.pdf/.png
              Figure2_Integrated_Dendrogram.pdf/.png
              Figure3_BioGroup_by_Origin.pdf/.png
  Companion: 04_functions.R
  Note    : BLAST must be run manually on NCBI before sections 6–10.
            See section 3b for instructions.
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

---

# 0. Configuration {.tabset}

Edit the paths and parameters in this block before running the notebook.

```{r config}
# ── Paths ─────────────────────────────────────────────────────────────────────
# All paths are relative to the project root.
# Set your working directory to the project root before knitting.
CASES_DIR <- "data/cases"
WGS_PATH  <- "data/wgs/ps_wgs_EcfliC.RDS"
BLAST_DIR <- "data/blast"
OUT_DIR   <- "results/04_best_vs_standard"

# BLAST JSON files must be placed in:
#   file.path(BLAST_DIR, "JSON")  e.g. blast_part_01.json, blast_part_02.json, …
BLAST_DIR_JSON <- file.path(BLAST_DIR, "JSON")

dir.create(OUT_DIR,          showWarnings = FALSE, recursive = TRUE)
dir.create(BLAST_DIR,        showWarnings = FALSE, recursive = TRUE)
dir.create(BLAST_DIR_JSON,   showWarnings = FALSE, recursive = TRUE)

# ── Pipeline parameters ───────────────────────────────────────────────────────
# Best pipeline (identified in step 03)
BEST_F      <- "Strict"      # Filtering  : maxEE = 0.5
BEST_ERR    <- "Binned"
BEST_POOL   <- "FALSE"
BEST_OMEGA  <- "1e-200"
BEST_BAND   <- "32"

# Standard pipeline (default DADA2 settings)
STD_F       <- "Loose"       # Filtering  : maxEE = 2
STD_ERR     <- "PacBio"
STD_POOL    <- "FALSE"
STD_OMEGA   <- "1e-40"
STD_BAND    <- "32"

# ── Pattern helpers ───────────────────────────────────────────────────────────
make_pattern <- function(f, err, pool, omega, band) {
  sprintf("F-%s__Err-%s__Pool-%s__O%s__B%s", f, err, pool, omega, band)
}

BEST_PATTERN <- make_pattern(BEST_F, BEST_ERR, BEST_POOL, BEST_OMEGA, BEST_BAND)
STD_PATTERN  <- make_pattern(STD_F,  STD_ERR,  STD_POOL,  STD_OMEGA,  STD_BAND)

cat("Best pattern    :", BEST_PATTERN, "\n")
cat("Standard pattern:", STD_PATTERN, "\n")
```

---

# 1. Load Libraries & Functions

```{r load-libs}
# ── Load the companion functions file ─────────────────────────────────────────
fn_file <- "04_functions.R"
if (!file.exists(fn_file))
  stop("Functions file not found: '", fn_file,
       "'\nPlace it in the same folder as this Rmd.")
source(fn_file)
load_libraries()
```

---

# 2. Load Input Data {.tabset}

## 2a. Locate Case Files

```{r locate-cases}
all_cases <- list.files(CASES_DIR, pattern = "\\.rds$", full.names = TRUE,
                        ignore.case = TRUE)
cat("Total case files found:", length(all_cases), "\n")

best_files <- all_cases[grepl(BEST_PATTERN, all_cases, fixed = TRUE)]
std_files  <- all_cases[grepl(STD_PATTERN,  all_cases, fixed = TRUE)]

cat("\nBest pipeline files (", length(best_files), "):\n")
cat(basename(best_files), sep = "\n")

cat("\nStandard pipeline files (", length(std_files), "):\n")
cat(basename(std_files), sep = "\n")

if (length(best_files) == 0) stop("No Best pipeline files found. Check CASES_DIR and BEST_PATTERN.")
if (length(std_files)  == 0) stop("No Standard pipeline files found. Check CASES_DIR and STD_PATTERN.")
```

## 2b. Load Phyloseq Objects

```{r load-ps}
ps_best_list <- lapply(best_files, load_case_ps)
ps_best      <- if (length(ps_best_list) == 1) ps_best_list[[1]] else
                  Reduce(merge_phyloseq, ps_best_list)

ps_std_list  <- lapply(std_files, load_case_ps)
ps_std       <- if (length(ps_std_list) == 1) ps_std_list[[1]] else
                  Reduce(merge_phyloseq, ps_std_list)

cat("Best phyloseq — taxa:", ntaxa(ps_best), "| samples:", nsamples(ps_best), "\n")
cat("Std  phyloseq — taxa:", ntaxa(ps_std),  "| samples:", nsamples(ps_std),  "\n")
```

## 2c. Load WGS Phyloseq

```{r load-wgs}
ps_wgs <- load_wgs_ps(WGS_PATH)
cat("WGS phyloseq — taxa:", ntaxa(ps_wgs), "| samples:", nsamples(ps_wgs), "\n")
```

---

# 3. Extract Sequences & Merge {.tabset}

## 3a. Extract ASV Sequences

```{r extract-seqs}
seqs_best <- get_asv_sequences(ps_best)
seqs_std  <- get_asv_sequences(ps_std)

# Prefix every ASV name with its pipeline label to guarantee uniqueness
# in the combined set and make pipeline membership unambiguous throughout.
names(seqs_best) <- paste0("Best_",     names(seqs_best))
names(seqs_std)  <- paste0("Standard_", names(seqs_std))

# Combine — no deduplication needed because names are now unique
seqs_all <- c(seqs_best, seqs_std)

best_asvs_canonical <- names(seqs_best)   # all "Best_*" entries
std_asvs_canonical  <- names(seqs_std)    # all "Standard_*" entries

cat("ASVs in Best     :", length(seqs_best), "\n")
cat("ASVs in Standard :", length(seqs_std),  "\n")
cat("Total ASVs       :", length(seqs_all),  "\n")

seqs_best_char <- as.character(seqs_best)
seqs_std_char  <- as.character(seqs_std)
n_shared <- sum(seqs_best_char %in% seqs_std_char)
cat("Shared sequences (same seq, in both pipelines):", n_shared, "\n")
cat("Best-only  :", sum(!seqs_best_char %in% seqs_std_char), "\n")
cat("Standard-only:", sum(!seqs_std_char %in% seqs_best_char), "\n")
```

## 3b. Write FASTA for BLAST

```{r write-fasta}
fasta_path    <- file.path(OUT_DIR, "04_all_asvs.fasta")
N_FASTA_PARTS <- 7

# Full FASTA
write_fasta(seqs = seqs_all, out_path = fasta_path)
cat("Full FASTA written to:", fasta_path, "\n\n")

# Split FASTA files — submit each part to NCBI BLAST web interface, then
# download "Download All > Single-file JSON2" and save into BLAST_DIR_JSON.
split_fasta_paths <- write_fasta(
  seqs     = seqs_all,
  out_path = file.path(BLAST_DIR, "blast_part.fasta"),
  n_parts  = N_FASTA_PARTS
)

cat("\nFASTA split files created:\n")
print(split_fasta_paths)

cat("\n⚠️  Submit each FASTA part to NCBI BLAST (https://blast.ncbi.nlm.nih.gov).\n")
cat("   Download results as: 'Download All > Single-file JSON2'\n")
cat("   Save files into BLAST_DIR_JSON as:\n")
for (i in seq_len(N_FASTA_PARTS)) {
  cat(sprintf("     blast_part_%02d.fasta  ->  %s/blast_part_%02d.json\n",
              i, BLAST_DIR_JSON, i))
}
cat("\nBLAST_DIR_JSON:", BLAST_DIR_JSON, "\n")
```

---

# 4. Edit Distance Matrix

Store the distance matrix and reuse it if present in the output folder.

```{r edit-distance, cache=FALSE}
# ── Distance matrix: load from cache or compute from scratch ──────────────────
dm_cache_path <- file.path(OUT_DIR, "cached_edit_distance_matrix.rds")

.dm_is_valid <- function(dm, expected_names) {
  if (is.null(dm)) return(FALSE)
  m <- tryCatch(as.matrix(dm), error = function(e) NULL)
  if (is.null(m)) return(FALSE)
  if (!all(expected_names %in% rownames(m))) return(FALSE)
  off_diag <- m[row(m) != col(m)]
  any(is.finite(off_diag))
}

force_recompute <- FALSE

if (file.exists(dm_cache_path)) {
  message(">>> Cache found — loading distance matrix from: ", dm_cache_path)
  cache_obj <- tryCatch(readRDS(dm_cache_path), error = function(e) NULL)

  if (!is.null(cache_obj) && .dm_is_valid(cache_obj$dm_bp, names(seqs_all))) {
    seqs_aligned <- cache_obj$seqs_aligned
    dm_bp        <- cache_obj$dm_bp
    message(">>> Cache is valid — using cached distance matrix.")
  } else {
    message(">>> Cache is STALE or INVALID — deleting and recomputing ...")
    file.remove(dm_cache_path)
    force_recompute <- TRUE
  }
} else {
  force_recompute <- TRUE
}

if (force_recompute) {

  cat("\n── Sequence composition diagnostics ──────────────────────────\n")
  cat("Best ASVs            :", length(seqs_best), "\n")
  cat("Standard ASVs        :", length(seqs_std),  "\n")
  cat("Total (with prefixes):", length(seqs_all),  "\n")

  n_dup_best <- sum(duplicated(seqs_best_char))
  n_dup_std  <- sum(duplicated(seqs_std_char))

  cat("Duplicated seqs within Best    :", n_dup_best,
      if (n_dup_best > 0) " <-- UNEXPECTED!" else " OK", "\n")
  cat("Duplicated seqs within Standard:", n_dup_std,
      if (n_dup_std > 0) " <-- UNEXPECTED!" else " OK", "\n")
  cat("──────────────────────────────────────────────────────────────\n\n")

  cat("Aligning", length(seqs_all), "sequences with DECIPHER...\n")
  seqs_aligned <- AlignSeqs(seqs_all, verbose = FALSE)

  cat("Computing edit distance matrix (pwalign Hamming)...\n")
  if (!requireNamespace("pwalign", quietly = TRUE))
    stop("pwalign required: BiocManager::install('pwalign')")
  dm_bp <- as.dist(as.matrix(pwalign::stringDist(seqs_aligned, method = "hamming")))

  # Validate
  dm_mat_check <- as.matrix(dm_bp)
  off_diag     <- dm_mat_check[row(dm_mat_check) != col(dm_mat_check)]

  cat("\n── Distance matrix diagnostics ───────────────────────────────\n")
  cat("Matrix dimensions     :", nrow(dm_mat_check), "×", ncol(dm_mat_check), "\n")
  cat("NA cells (off-diag)   :", sum(is.na(off_diag)), "\n")
  cat("Distance range (bp)   :", range(off_diag, na.rm = TRUE), "\n")
  cat("──────────────────────────────────────────────────────────────\n\n")

  if (all(is.na(off_diag)))
    stop("FATAL: distance matrix is entirely NA/NaN.")

  message(">>> Saving distance matrix to cache: ", dm_cache_path)
  saveRDS(list(seqs_aligned = seqs_aligned, dm_bp = dm_bp), file = dm_cache_path)
  message(">>> Distance matrix cached successfully.")
}

dm_mat_final   <- as.matrix(dm_bp)
off_diag_final <- dm_mat_final[row(dm_mat_final) != col(dm_mat_final)]
cat("Distance matrix ready. Dimensions:", nrow(dm_mat_final), "×", ncol(dm_mat_final), "\n")
cat("Distance range (off-diagonal):",
    range(off_diag_final, na.rm = TRUE)[1], "to",
    range(off_diag_final, na.rm = TRUE)[2], "bp\n")
```

---

# 5. FIGURE 1 — Edit Distance Distributions

Three panels showing within-Best, within-Standard, and cross-pipeline edit
distances to the closest matching sequence in each comparison group.

```{r figure1, fig.width=12, fig.height=10}
best_asvs <- intersect(best_asvs_canonical, rownames(as.matrix(dm_bp)))
std_asvs  <- intersect(std_asvs_canonical,  rownames(as.matrix(dm_bp)))

cat("Best ASVs in dm_bp     :", length(best_asvs), "\n")
cat("Standard ASVs in dm_bp :", length(std_asvs),  "\n")
cat("Name overlap (must = 0):", length(intersect(best_asvs, std_asvs)), "\n")

if (length(best_asvs) == 0) stop("No Best ASVs in dm_bp.")
if (length(std_asvs)  == 0) stop("No Standard ASVs in dm_bp.")

p1 <- figure1_edit_dist(
  dm_bp     = dm_bp,
  best_asvs = best_asvs,
  std_asvs  = std_asvs,
  colors    = list(best  = "#4472C4",
                   std   = "#70AD47",
                   cross = "#ED7D31")
)
print(p1)
```

```{r save-figure1}
fig1_path <- file.path(OUT_DIR, "Figure1_Edit_Distance_Distribution.pdf")
ggsave(fig1_path, plot = p1, width = 12, height = 10, device = "pdf")
cat("Figure 1 saved:", fig1_path, "\n")

fig1_png  <- file.path(OUT_DIR, "Figure1_Edit_Distance_Distribution.png")
ggsave(fig1_png,  plot = p1, width = 12, height = 10, dpi = 300, device = "png")
cat("Figure 1 PNG saved:", fig1_png, "\n")
```

---

# 6. BLAST Loading & LCA

> ⚠️ **Run BLAST externally on the NCBI website**, then place the JSON results
> into `data/blast/JSON/` and re-run from this section.
>
> **Required downloads (for each FASTA part):**
>
> 1. **Single-file JSON** → save as `data/blast/JSON/blast_part_NN.json`
>    *(provides: query title, query length, accession, taxid, sciname, title,*
>    *identity, align_len, gaps, query_from, query_to, hit_from, hit_to,*
>    *evalue, bit_score for each HSP)*
>
> **query_coverage** is computed directly from the JSON coordinates:
> `(query_to - query_from + 1) / query_len * 100`
>
> All 7 JSON files are read automatically from `BLAST_DIR_JSON`.

```{r blast-load}
blast_df <- NULL
lca_df   <- NULL

# ── Cache: blast_df_parsed.csv ────────────────────────────────────────────────
# On first run, all JSON files are parsed and the result is saved as
# blast_df_parsed.csv inside BLAST_DIR for fast re-runs.
# Delete that CSV to force a re-parse (e.g. after updating JSON files).
blast_cache_path <- file.path(BLAST_DIR, "blast_df_parsed.csv")

if (file.exists(blast_cache_path)) {

  message(">>> blast_df_parsed.csv found — loading cached BLAST results.")
  message(">>> Path: ", blast_cache_path)
  blast_df <- read.csv(blast_cache_path, header = TRUE,
                       stringsAsFactors = FALSE, na.strings = c("", "NA"))
  # Restore numeric types
  if ("pident"      %in% colnames(blast_df)) blast_df$pident      <- as.numeric(blast_df$pident)
  if ("query_cover" %in% colnames(blast_df)) blast_df$query_cover <- as.numeric(blast_df$query_cover)
  if ("evalue"      %in% colnames(blast_df)) blast_df$evalue      <- as.numeric(blast_df$evalue)
  if ("bitscore"    %in% colnames(blast_df)) blast_df$bitscore    <- as.numeric(blast_df$bitscore)

  message(">>> Loaded: ", nrow(blast_df), " rows | ",
          dplyr::n_distinct(blast_df$qseqid), " unique ASVs")
  qcov_ok <- "query_cover" %in% colnames(blast_df) && !all(is.na(blast_df$query_cover))
  message(">>> query_cover available: ", if (qcov_ok) "YES" else "NO (all NA)")

} else {

  message(">>> blast_df_parsed.csv not found — running JSON parser.")
  message(">>> Looking for JSON files in: ", BLAST_DIR_JSON)

  json_present <- dir.exists(BLAST_DIR_JSON) &&
                  length(list.files(BLAST_DIR_JSON, pattern = "\\.json$")) > 0

  if (!json_present) {
    cat("No JSON files found in:", BLAST_DIR_JSON, "\n")
    cat("Expected directory structure:\n")
    cat("  ", BLAST_DIR_JSON, "/blast_part_01.json  … blast_part_NN.json\n")
    blast_df <- NULL
    lca_df   <- NULL
  } else {
    blast_df <- load_blast_results(BLAST_DIR)
    write.csv(blast_df, blast_cache_path, row.names = FALSE)
    message(">>> Saved blast_df_parsed.csv: ", nrow(blast_df), " rows | ",
            ncol(blast_df), " columns")
    message(">>> Path: ", blast_cache_path)
  }
}
```

```{r lca}
# ── LCA: genus-level consensus from top BLAST hits ────────────────────────────
# top_n hits filtered by min_pident; genus-level consensus as LCA.
# All queries from blast_df are preserved (n_hits = 0 if none pass filter).
if (!is.null(blast_df)) {
  lca_df <- compute_lca(blast_df, top_n = 10, min_pident = 97)

  cat("── LCA Summary ─────────────────────────────────────────────────\n")
  cat("Columns in lca_df      :", paste(colnames(lca_df), collapse = ", "), "\n")
  cat("ASVs in lca_df         :", nrow(lca_df), "\n")
  cat("Missing lca_name       :", sum(is.na(lca_df$lca_name)), "\n")
  cat("────────────────────────────────────────────────────────────────\n")
  print(head(lca_df))

} else {
  lca_df <- data.frame(qseqid = character(), lca_name = character(),
                       n_hits = integer(), stringsAsFactors = FALSE)
}
```

---

# 7. Trust Tier Assignment

Each ASV is classified into one of five mutually exclusive tiers based on its
BLAST hits against NCBI. A **perfect hit** is defined as 100% identity AND
100% query coverage.

- **Tier 1 — All perfect E. coli (or WGS match):** Every BLAST hit has 100% identity and 100% coverage with *Escherichia coli*, OR the ASV has an exact sequence match (edit distance = 0) to the local WGS database.
- **Tier 2 — At least one perfect E. coli hit:** At least one BLAST hit is 100% identity + 100% coverage *E. coli*, but not all hits satisfy this criterion.
- **Tier 3 — No perfect hits, all hits E. coli:** No perfect hits, but every BLAST hit is related to *Escherichia coli*.
- **Tier 4 — No perfect hits, mixed taxonomy:** No perfect hits, and BLAST hits are a mixture of *E. coli* and non-*E. coli* organisms.
- **Tier 5 — No E. coli hits:** No BLAST hit is related to *Escherichia coli*.

Dendrogram grouping: **E. coli** = Tiers 1 + 2 | **Ambiguous** = Tiers 3 + 4 | **Non E. coli** = Tier 5.

```{r trust-tiers}
if (!is.null(blast_df) && nrow(lca_df) > 0) {

  # ── Build base ASV data frame ──────────────────────────────────────────────
  asv_base <- data.frame(asv = names(seqs_all), stringsAsFactors = FALSE)

  # top_pident: max BLAST %identity per ASV (kept in asv_base, NOT in lca_df)
  top_pident_df <- blast_df %>%
    group_by(qseqid) %>%
    summarise(top_pident = max(pident, na.rm = TRUE), .groups = "drop")

  asv_base <- asv_base %>%
    left_join(top_pident_df, by = c("asv" = "qseqid")) %>%
    mutate(top_pident = replace_na(top_pident, 0))

  # top_qcov: max query coverage per ASV
  if ("query_cover" %in% colnames(blast_df) && !all(is.na(blast_df$query_cover))) {
    top_qcov_df <- blast_df %>%
      mutate(qcov_val = as.numeric(query_cover)) %>%
      group_by(qseqid) %>%
      summarise(top_qcov = max(qcov_val, na.rm = TRUE), .groups = "drop")
    asv_base <- asv_base %>%
      left_join(top_qcov_df, by = c("asv" = "qseqid")) %>%
      mutate(top_qcov = replace_na(top_qcov, 0))
    cat("top_qcov: range", round(min(top_qcov_df$top_qcov), 1), "–",
        round(max(top_qcov_df$top_qcov), 1), "\n")
  } else {
    asv_base$top_qcov <- 0
    cat("NOTE: query_cover not available — top_qcov = 0 for all ASVs.\n")
  }

  # ── WGS confirmation via edit distance = 0 ────────────────────────────────
  # An ASV is "WGS confirmed" if its edit distance to any WGS sequence is 0.
  seqs_wgs      <- get_asv_sequences(ps_wgs)
  seqs_combined <- c(seqs_all, seqs_wgs)
  seqs_combined <- seqs_combined[!duplicated(names(seqs_combined))]

  cat("Computing WGS-confirmation distances ...\n")
  cat("  Amplicon ASVs (prefixed):", length(seqs_all), "\n")
  cat("  WGS ASVs                :", length(seqs_wgs), "\n")

  seqs_combined_aligned <- AlignSeqs(seqs_combined, verbose = FALSE)
  if (!requireNamespace("pwalign", quietly = TRUE))
    stop("pwalign required: BiocManager::install('pwalign')")
  dm_combined_bp <- as.matrix(
    pwalign::stringDist(seqs_combined_aligned, method = "hamming")
  )
  rownames(dm_combined_bp) <- names(seqs_combined_aligned)
  colnames(dm_combined_bp) <- names(seqs_combined_aligned)

  amp_ids <- names(seqs_all)
  wgs_ids <- names(seqs_wgs)

  amp_ids_in <- intersect(amp_ids, rownames(dm_combined_bp))
  wgs_ids_in <- intersect(wgs_ids, colnames(dm_combined_bp))

  if (length(wgs_ids_in) == 0) {
    warning("No WGS ASV sequences found in combined distance matrix.")
    wgs_confirmed_asvs <- character(0)
  } else {
    dm_cross           <- dm_combined_bp[amp_ids_in, wgs_ids_in, drop = FALSE]
    min_cross_dist     <- apply(dm_cross, 1, min, na.rm = TRUE)
    wgs_confirmed_asvs <- names(min_cross_dist)[min_cross_dist == 0]
  }

  cat("  WGS-confirmed ASVs (edit dist = 0):", length(wgs_confirmed_asvs), "\n")
  cat("  Non-confirmed ASVs               :", length(amp_ids) - length(wgs_confirmed_asvs), "\n")

  asv_base <- asv_base %>%
    mutate(wgs_confirmed = asv %in% wgs_confirmed_asvs)

  # ── Input validation ───────────────────────────────────────────────────────
  required_asv_cols <- c("asv", "top_pident", "top_qcov", "wgs_confirmed")
  missing_asv <- setdiff(required_asv_cols, colnames(asv_base))
  if (length(missing_asv) > 0)
    stop("trust-tier: asv_base missing: ", paste(missing_asv, collapse = ", "))

  required_lca_cols <- c("qseqid", "lca_name", "n_hits")
  missing_lca <- setdiff(required_lca_cols, colnames(lca_df))
  if (length(missing_lca) > 0)
    stop("trust-tier: lca_df missing: ", paste(missing_lca, collapse = ", "))

  if ("top_pident" %in% colnames(lca_df)) {
    warning("lca_df contains 'top_pident' — removing to prevent join collision.")
    lca_df <- dplyr::select(lca_df, -top_pident)
  }

  # ── Assign tiers ───────────────────────────────────────────────────────────
  trust_df <- assign_trust_tiers(
    asv_df       = asv_base,
    wgs_ps       = ps_wgs,
    lca_df       = lca_df,
    blast_df     = blast_df,
    target_genus = "Escherichia"
  )

  cat("\n── Trust Tier Distribution (5-tier system) ─────────────────────\n")
  cat("By individual tier:\n")
  print(table(trust_df$trust_label))
  cat("\nBy biological group (E. coli / Ambiguous / Non E. coli):\n")
  print(table(trust_df$trust_group))
  cat("\nWGS-confirmed breakdown:\n")
  print(table(WGS = trust_df$wgs_confirmed, Tier = trust_df$trust_tier))
  cat("────────────────────────────────────────────────────────────────\n")

} else {
  trust_df <- data.frame(
    asv           = names(seqs_all),
    trust_label   = "Tier 5: No E. coli hits",
    trust_tier    = 5L,
    trust_group   = "Non E. coli",
    wgs_confirmed = FALSE,
    stringsAsFactors = FALSE
  )
  cat("No BLAST data: all ASVs assigned Tier 5 (No E. coli hits) as placeholder.\n")
}
```

---

# 8. ASV Metadata for Figure 2

```{r asv-meta}
asv_meta <- build_asv_meta(
  ps_best        = ps_best,
  ps_std         = ps_std,
  trust_df       = trust_df,
  best_asvs_can  = best_asvs_canonical,
  std_asvs_can   = std_asvs_canonical,
  seqs_best_char = seqs_best_char,
  seqs_std_char  = seqs_std_char
)

cat("ASV metadata rows:", nrow(asv_meta), "\n")
cat("Origin counts:\n");      print(table(asv_meta$origin))
cat("Trust tier counts:\n");  print(table(asv_meta$trust_label))
cat("Trust group counts:\n"); print(table(asv_meta$trust_group))
```

---

# 9. FIGURE 2 — Integrated Dendrogram

A **single, integrated figure** combining:

- Hierarchical clustering dendrogram (edit-distance based, NOT phylogenetic inference)
- Trust tier strip (5 individual tiers)
- Biological group strip (E. coli / Ambiguous / Non E. coli)
- Abundance heatmap (% relative abundance, sqrt-scaled)
- Prevalence annotation
- Sequence alignment (nucleotide level)
- Origin annotation (Best / Standard / Shared)

```{r figure2-prep}
ps_merged <- merge_phyloseq(ps_best, ps_std)
ps_rel    <- transform_sample_counts(ps_merged, function(x) x / sum(x))
otu_mat   <- as(otu_table(ps_rel), "matrix")
if (!taxa_are_rows(ps_rel)) otu_mat <- t(otu_mat)

asvs_in_dm   <- rownames(as.matrix(dm_bp))
strip_prefix <- function(x) sub("^(Best_|Standard_)", "", x)
orig_names   <- strip_prefix(asvs_in_dm)

rows_found  <- orig_names %in% rownames(otu_mat)
otu_sub_raw <- otu_mat[orig_names[rows_found], , drop = FALSE]
rownames(otu_sub_raw) <- asvs_in_dm[rows_found]

missing_prefixed <- asvs_in_dm[!rows_found]
if (length(missing_prefixed) > 0) {
  zero_rows <- matrix(0, nrow = length(missing_prefixed), ncol = ncol(otu_sub_raw),
                      dimnames = list(missing_prefixed, colnames(otu_sub_raw)))
  otu_sub <- rbind(otu_sub_raw, zero_rows)[asvs_in_dm, , drop = FALSE]
} else {
  otu_sub <- otu_sub_raw[asvs_in_dm, , drop = FALSE]
}

asv_meta_sub <- asv_meta %>% filter(asv %in% asvs_in_dm)
dm_sub       <- as.dist(as.matrix(dm_bp)[asvs_in_dm, asvs_in_dm])

cat("ASVs entering Figure 2:", length(asvs_in_dm), "\n")
```

```{r figure2, fig.width=28, fig.height=20}
p2 <- figure2_dendrogram(
  dm_bp        = dm_sub,
  asv_meta     = asv_meta_sub,
  seqs_aligned = seqs_aligned[names(seqs_aligned) %in% asvs_in_dm],
  sample_abund = otu_sub,
  merge_dist   = 0
)
print(p2)
```

```{r save-figure2}
fig2_path <- file.path(OUT_DIR, "Figure2_Integrated_Dendrogram.pdf")
ggsave(fig2_path, plot = p2, width = 28, height = 20, device = "pdf", limitsize = FALSE)
cat("Figure 2 saved:", fig2_path, "\n")

fig2_png <- file.path(OUT_DIR, "Figure2_Integrated_Dendrogram.png")
ggsave(fig2_png, plot = p2, width = 28, height = 20, dpi = 200, device = "png", limitsize = FALSE)
cat("Figure 2 PNG saved:", fig2_png, "\n")
```

---

# 10. FIGURE 3 — Bio Group Distribution by ASV Origin

This section compares the **Bio Group composition** (*E. coli* / Ambiguous /
Non-*E. coli*) across the three ASV categories defined by pipeline overlap:

- **Shared** — ASVs present in both the Best and Standard pipelines
- **Unique to Best** — ASVs detected only by the strict Best pipeline
- **Unique to Standard** — ASVs detected only by the permissive Standard pipeline

Each bar shows the proportion (top panel) and absolute count (bottom panel) of
Bio Groups within each category, using the same classification from the dendrogram.

```{r figure3, fig.width=9, fig.height=10}
p3 <- figure3_biogroup_by_origin(asv_meta)
print(p3)
```

```{r save-figure3}
fig3_pdf <- file.path(OUT_DIR, "Figure3_BioGroup_by_Origin.pdf")
ggsave(fig3_pdf, plot = p3, width = 9, height = 10, device = "pdf")
cat("Figure 3 PDF saved:", fig3_pdf, "\n")

fig3_png <- file.path(OUT_DIR, "Figure3_BioGroup_by_Origin.png")
ggsave(fig3_png, plot = p3, width = 9, height = 10, dpi = 300, device = "png")
cat("Figure 3 PNG saved:", fig3_png, "\n")
```

---

# 11. Session Info

```{r session-info}
sessionInfo()
```
