---
title: "06 — Longitudinal E. coli ASV Carriage"
author: "Andrés Catalán Tatay"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

<!--
================================================================================
  Project : fliC Amplicon Sequencing for High-Resolution Profiling of E. coli
            in the Infant Gut Microbiome
  Author  : Andrés Catalán Tatay
  Part    : Analysis — Step 06
  Purpose : Longitudinal E. coli ASV carriage profiling across repeated faecal
            samples from the same individual. Compares detection patterns
            between the BEST and STANDARD DADA2 pipelines.
            Only ASVs with perfect BLAST support (100% identity, 100% query
            coverage, Escherichia coli annotation) are retained.
  Inputs  : data/cases/case_049__F-Strict__Err-Binned__Pool-FALSE__O1e-200__B32.rds
             data/cases/case_031__F-Loose__Err-PacBio__Pool-FALSE__O1e-40__B32.rds
             data/wgs/phyloseq_NCBInt_process_all_sub_OMEGA_A-200_poolFALSE_minQ30EE05.RDS
             results/04_best_vs_standard/blast/blast_df_parsed.csv
  Outputs : results/06_longitudinal_carriage/
              longitudinal_carriage_<subject>_<pipeline>.png/.pdf
              06_carriage_df.csv
              06_carriage_summary.csv
  Companion: 06_Longitudinal_Carriage_functions.R
  Note    : Requires BLAST results from step 04 to be completed first.
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width  = 10,
  fig.height = 6
)
```

---

## Conceptual background

This analysis evaluates whether *E. coli* ASVs inferred by *fliC* PacBio amplicon
sequencing show longitudinal carriage patterns across repeated samples from the same
individual.

By comparing the **BEST** and **STANDARD** DADA2 pipelines, the analysis assesses
whether pipeline choice affects the detection and persistence of *E. coli* ASVs over
time.

Only ASVs with **perfect BLAST support** for *Escherichia coli* are retained:

- 100% identity (`pident == 100`)
- 100% query coverage (`query_cover == 100`)
- *Escherichia coli* annotation in the BLAST `sscinames` field

Each plot shows **filled circles only**, sized by relative abundance at that timepoint.
Only timepoints with a true detection (Abundance > 0) are plotted — absent timepoints
leave an empty space on the x-axis.

---

## 1. Setup

### 1.1 Libraries

```{r libraries}
library(tidyverse)
library(phyloseq)
library(ggplot2)
```

### 1.2 Source helper functions

```{r source-functions}
# ── Load the companion functions file ─────────────────────────────────────────
fn_file <- "06_Longitudinal_Carriage_functions.R"
if (!file.exists(fn_file))
  stop("Functions file not found: '", fn_file,
       "'\nPlace it in the same folder as this Rmd.")
source(fn_file)
```

---

## 2. Paths

```{r paths}
# ── Paths ─────────────────────────────────────────────────────────────────────
# All paths are relative to the project root.
# Set your working directory to the project root before knitting.

cases_dir <- "data/cases"

case_path_BEST <- file.path(
  cases_dir,
  "case_049__F-Strict__Err-Binned__Pool-FALSE__O1e-200__B32.rds"
)

case_path_STANDARD <- file.path(
  cases_dir,
  "case_031__F-Loose__Err-PacBio__Pool-FALSE__O1e-40__B32.rds"
)

# Longitudinal metadata source (saved with save(), not saveRDS())
florentin_phyloseq_path <-
  "data/wgs/phyloseq_NCBInt_process_all_sub_OMEGA_A-200_poolFALSE_minQ30EE05.RDS"

# BLAST results from step 04
blast_file <-
  "results/04_best_vs_standard/blast/blast_df_parsed.csv"

output_dir <- "results/06_longitudinal_carriage"

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

---

## 3. Load data

### 3.1 BEST and STANDARD phyloseq objects

```{r load-cases}
message("Loading BEST case...")
ps_best_raw <- load_case_phyloseq(case_path_BEST)

message("Loading STANDARD case...")
ps_standard_raw <- load_case_phyloseq(case_path_STANDARD)

cat("BEST     — taxa:", ntaxa(ps_best_raw),     "| samples:", nsamples(ps_best_raw),     "\n")
cat("STANDARD — taxa:", ntaxa(ps_standard_raw), "| samples:", nsamples(ps_standard_raw), "\n")
```

### 3.2 Longitudinal metadata source

```{r load-florentin}
message("Loading longitudinal metadata phyloseq (via load())...")
ps_florentin <- load_florentin_phyloseq(florentin_phyloseq_path)
cat("Metadata source — taxa:", ntaxa(ps_florentin), "| samples:", nsamples(ps_florentin), "\n")
```

---

## 4. Attach longitudinal metadata

```{r attach-metadata}
meta_cols <- c("Subject", "days_since_baseline", "weeks_since_baseline",
               "Type", "Type2", "n_sample", "Time")

message("\nAttaching metadata to BEST...")
ps_best <- attach_longitudinal_metadata(ps_best_raw, ps_florentin, meta_cols)

message("\nAttaching metadata to STANDARD...")
ps_standard <- attach_longitudinal_metadata(ps_standard_raw, ps_florentin, meta_cols)
```

---

## 5. Filter to individual samples only

```{r filter-individual}
message("\nFiltering BEST to individual samples...")
ps_best <- filter_individual_samples(ps_best)

message("\nFiltering STANDARD to individual samples...")
ps_standard <- filter_individual_samples(ps_standard)

cat("\nAfter filtering:\n")
cat("  BEST     — samples:", nsamples(ps_best),     "\n")
cat("  STANDARD — samples:", nsamples(ps_standard), "\n")
```

---

## 6. Parse BLAST results and define E. coli ASVs

```{r parse-blast}
message("\nParsing BLAST file...")
ecoli_asvs <- parse_ecoli_asvs(blast_file)

ecoli_asvs %>%
  count(Pipeline) %>%
  knitr::kable(caption = "E. coli ASV counts by pipeline (perfect BLAST hits only)")
```

---

## 7. Determine subjects

```{r get-subjects}
subjects_best <- as(sample_data(ps_best), "data.frame") %>%
  filter(!is.na(Subject)) %>%
  distinct(Subject) %>%
  pull(Subject) %>%
  as.character()

subjects_standard <- as(sample_data(ps_standard), "data.frame") %>%
  filter(!is.na(Subject)) %>%
  distinct(Subject) %>%
  pull(Subject) %>%
  as.character()

subjects <- union(subjects_best, subjects_standard)
cat("Subjects to analyse:", paste(subjects, collapse = ", "), "\n")
```

---

## 8. Run longitudinal carriage analysis

```{r run-analysis}
results <- run_longitudinal_carriage(
  ps_best      = ps_best,
  ps_standard  = ps_standard,
  ecoli_asvs   = ecoli_asvs,
  subjects     = subjects,
  output_dir   = output_dir,
  group_by     = "days_since_baseline",

  # Abundance size mapping
  abund_breaks = c(-Inf, 0.5, 1, 2, 5, Inf),
  abund_labels = c("< 0.5%", "0.5-1%", "1-2%", "2-5%", "> 5%"),
  abund_sizes  = c("< 0.5%" = 1.0, "0.5-1%" = 2.5,
                   "1-2%"   = 4.0, "2-5%"   = 6.0, "> 5%" = 8.0),

  # All prevalence thresholds = 0 (keep all E. coli ASVs)
  prev_high_cutoff    = 0,
  prev_high_min_ab    = 0,
  prev_med_cutoff     = 0,
  prev_med_min_ab     = 0,
  prev_low_min_ab     = 0,
  min_subject_samples = 0,
  peak_trust_abund    = 0,

  fig_width  = 10,
  fig_height = 6
)
```

---

## 9. Display plots

```{r display-plots, results='asis', fig.width=10, fig.height=6}
for (plot_name in names(results$individual_plots)) {
  cat("\n\n### ", gsub("_", " — ", plot_name), "\n\n")
  print(results$individual_plots[[plot_name]])
}
```

---

## 10. Carriage dataframe

```{r carriage-df}
# Master long-format dataframe of all detections across subjects and pipelines.
# Saved automatically to output_dir as 06_carriage_df.csv.
results$carriage_df %>%
  head(20) %>%
  knitr::kable(caption = "First 20 rows of the carriage dataframe")
```

---

## 11. Summary table

```{r summary-table}
knitr::kable(
  results$summary_df,
  caption = "E. coli ASV carriage summary per Subject and Pipeline"
)
```

---

## 12. Session info

```{r session-info}
sessionInfo()
```
