---
title: "03 — DADA2 Parameter Benchmarking: Metric Modelling & Parameter Selection"
author: "Andrés Catalán Tatay"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

<!--
================================================================================
  Project : fliC Amplicon Sequencing for High-Resolution Profiling of E. coli
            in the Infant Gut Microbiome
  Author  : Andrés Catalán Tatay
  Part    : Analysis — Step 03
  Purpose : Second-stage benchmarking of the DADA2 parameter grid.
            Part 1: models the effect of each parameter on mock community
            quality metrics (chimeric reads, ASV accuracy, read recovery).
            Part 2: ranks all 96 parameter combinations using a composite
            score and identifies the optimal pipeline configuration.
  Inputs  : data/mock/mock_trimmed_4.fasta  — theoretical mock reference
             data/cases/                    — 96 RDS files (DADA2 outputs)
  Outputs : results/03_metric_modelling/
              tables/         — metrics, model coefficients, marginal means,
                                ranking tables
              figures/        — marginal means, boxplots, coefficient plots,
                                heatmap overview, ranking scatter and bar plots
              model_summaries/ — plain-text model summaries per metric
  Companion: 03_case_metric_modeling_functions.R
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 11,
  fig.height= 7,
  dpi       = 150,
  out.width = "100%"
)
```

---

# Overview

This document performs a **second-stage benchmarking analysis** of the DADA2
parameter grid. It **does not** repeat the PERMANOVA / community-structure
analysis from step 02. Instead, it focuses on:

**PART 1 — Parameter effects on benchmark metrics**
Using only the *mockmix* samples from all 96 cases, we compute interpretable
quality metrics and model the effect of each DADA2 parameter.

**PART 2 — Best parameter combination selection**
We rank all 96 parameter combinations according to two biologically meaningful
decision criteria and identify the top 3 score levels.

---

# Setup

## Load packages

```{r load-packages}
# Core packages required — install if missing:
# install.packages(c("ggplot2","dplyr","tidyr","vegan"))

library(ggplot2)
library(dplyr)
library(tidyr)
library(vegan)   # required for rrarefy() used in load_all_cases()

# Check that required packages are available
required_pkgs <- c("ggplot2", "dplyr", "tidyr", "vegan")
missing_pkgs  <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Please install missing packages: ", paste(missing_pkgs, collapse = ", "))
}

cat("All required packages loaded.\n")
```

## Source helper functions

```{r source-functions}
# ── Load the companion functions file ─────────────────────────────────────────
# Place 03_case_metric_modeling_functions.R in the same folder as this Rmd.
fn_file <- "03_case_metric_modeling_functions.R"
if (!file.exists(fn_file))
  stop("Functions file not found: '", fn_file,
       "'\nPlace it in the same folder as this Rmd.")
source(fn_file)
cat("Functions loaded from:", fn_file, "\n")
```

## Define paths

```{r define-paths}
# ── Paths ─────────────────────────────────────────────────────────────────────
# All paths are relative to the project root.
# Set your working directory to the project root before knitting.

# INPUT
MOCK_FASTA_PATH <- "data/mock/mock_trimmed_4.fasta"
CASES_DIR       <- "data/cases"

# OUTPUT ROOT
OUTPUT_ROOT <- "results/03_metric_modelling"

# OUTPUT SUB-DIRECTORIES
DIR_TABLES  <- file.path(OUTPUT_ROOT, "tables")
DIR_FIGURES <- file.path(OUTPUT_ROOT, "figures")
DIR_MODELS  <- file.path(OUTPUT_ROOT, "model_summaries")

# Create all output directories
for (d in c(DIR_TABLES, DIR_FIGURES, DIR_MODELS)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

cat("Output directories ready under:\n", OUTPUT_ROOT, "\n")
```

---

# Load data

## Load theoretical mock reference sequences

```{r load-mock}
# Parse the FASTA file. The function handles IUPAC ambiguity codes (e.g. V in
# ESBL15) — matching is performed with full IUPAC awareness (see functions file).
mock_seqs <- parse_fasta(MOCK_FASTA_PATH)

cat("Loaded", length(mock_seqs), "theoretical mock sequences:\n")
for (nm in names(mock_seqs)) {
  cat(" -", nm, ":", nchar(mock_seqs[nm]), "bp")
  ambig <- setdiff(unique(strsplit(mock_seqs[nm], "")[[1]]), c("A","C","G","T"))
  if (length(ambig) > 0) cat("  [IUPAC ambiguities:", paste(ambig, collapse=","), "]")
  cat("\n")
}
```

## Load all 96 case files

```{r load-cases}
# Scans CASES_DIR for all files matching ^case_.*\.rds$
# Files that fail to load are silently skipped with a warning.
cases_list <- load_all_cases(CASES_DIR)

cat("\nTotal cases successfully loaded:", length(cases_list), "\n")
```

---

# PART 1 — Effect of parameters on benchmark metrics

## Compute metrics for all cases

```{r compute-metrics}
# For each case, metrics are computed on mockmix samples only.
# See the functions file for precise definitions of each metric.
cat("Computing metrics for all cases...\n")
metrics_df <- build_metrics_table(cases_list, mock_seqs)

# Apply sensible factor level ordering for plotting
metrics_df <- factorise_parameters(metrics_df)

cat("\nMetrics table dimensions:", nrow(metrics_df), "rows ×", ncol(metrics_df), "cols\n")
cat("Parameters found:\n")
cat("  filter_mode :", paste(levels(metrics_df$filter_mode),  collapse = ", "), "\n")
cat("  error_model :", paste(levels(metrics_df$error_model),  collapse = ", "), "\n")
cat("  pool_mode   :", paste(levels(metrics_df$pool_mode),    collapse = ", "), "\n")
cat("  omega_raw   :", paste(levels(metrics_df$omega_raw),    collapse = ", "), "\n")
cat("  band_size   :", paste(levels(metrics_df$band_size),    collapse = ", "), "\n")
```

## Preview metrics table

```{r preview-metrics}
# Show all 6 metrics for the first few cases
metrics_df |>
  select(case_id, filter_mode, error_model, pool_mode, omega_raw, band_size,
         pct_chimeric_reads, mean_reads_per_sample, total_asvs,
         pct_asvs_matching_mock, pct_asvs_not_matching_mock,
         pct_reads_matching_mock) |>
  head(10)
```

## Save metrics table

```{r save-metrics}
save_csv(metrics_df, file.path(DIR_TABLES, "01_case_metrics.csv"))
```

## Summary statistics per metric

```{r summary-stats}
target_metrics <- c(
  "pct_chimeric_reads",
  "mean_reads_per_sample",
  "total_asvs",
  "pct_asvs_matching_mock",
  "pct_asvs_not_matching_mock",
  "pct_reads_matching_mock"
)

summary_stats <- do.call(rbind, lapply(target_metrics, function(m) {
  x <- metrics_df[[m]]
  data.frame(
    metric = m,
    mean   = round(mean(x, na.rm=TRUE), 3),
    sd     = round(sd(x,   na.rm=TRUE), 3),
    min    = round(min(x,  na.rm=TRUE), 3),
    median = round(median(x,na.rm=TRUE),3),
    max    = round(max(x,  na.rm=TRUE), 3),
    n_na   = sum(is.na(x))
  )
}))

print(summary_stats)
save_csv(summary_stats, file.path(DIR_TABLES, "02_metric_summary_stats.csv"))
```

---

## Statistical modelling

We fit one regression model per metric, treating all five DADA2 parameters as
fixed categorical predictors.

**Model choices (see functions file for full rationale):**

| Metric | Model | Link |
|---|---|---|
| % chimeric reads | Quasibinomial GLM | logit |
| Mean reads/sample | OLS | identity (on log scale) |
| Total ASVs | Poisson GLM | log |
| % ASVs matching mock | Quasibinomial GLM | logit |

```{r fit-models}
cat("Fitting regression models...\n")
models_list <- fit_all_models(metrics_df)

# Print summaries to console and save to disk
for (metric in names(models_list)) {
  m <- models_list[[metric]]
  if (is.null(m)) next
  cat("\n", paste(rep("=", 60), collapse=""), "\n")
  cat("METRIC:", metric, "\n")
  cat("FAMILY:", m$family_used, "\n")
  cat(paste(rep("-", 60), collapse=""), "\n")
  print(m$summary)
  save_model_summary(m, file.path(DIR_MODELS, paste0(metric, "_model_summary.txt")))
}
```

## Coefficient table (all models combined)

```{r coef-table}
all_coefs <- extract_all_coefficients(models_list)
print(all_coefs)
save_csv(all_coefs, file.path(DIR_TABLES, "03_all_model_coefficients.csv"))
```

## Marginal means (observed group means per parameter level)

```{r marginal-means}
mm_df <- all_marginal_means(metrics_df)
save_csv(mm_df, file.path(DIR_TABLES, "04_marginal_means.csv"))
head(mm_df, 20)
```

---

## Figures — Part 1

### Figure 1: Marginal means plots (one per metric)

For each metric we show the observed mean ± 95% CI across all parameter levels.
This is the clearest way to communicate parameter effects without model
back-transformation.

```{r fig-marginal-means, fig.height=5}
metric_labels <- c(
  pct_chimeric_reads     = "% chimeric reads in mock (lower = better)",
  mean_reads_per_sample  = "Mean reads per sample (mock)",
  pct_asvs_matching_mock = "% ASVs matching theoretical mock (higher = better)",
  total_asvs             = "Total observed ASVs in mock samples"
)

for (m in names(metric_labels)) {
  p <- plot_marginal_means(mm_df, m, y_label = metric_labels[m])
  print(p)
  save_plot(p,
            file.path(DIR_FIGURES, paste0("01_marginal_means_", m, ".png")),
            width = 14, height = 5)
}
```

### Figure 2: Boxplots per parameter × metric

For each combination of metric and parameter, show the distribution of values
across cases — this reveals not just the mean but the spread and any outliers.

```{r fig-boxplots, fig.height=5, fig.width=8}
params <- c("filter_mode","error_model","pool_mode","omega_raw","band_size")
param_labels <- c(
  filter_mode  = "Filter mode",
  error_model  = "Error model",
  pool_mode    = "Pooling mode",
  omega_raw    = "Omega (ω)",
  band_size    = "Band size"
)

for (m in names(metric_labels)) {
  for (param in params) {
    p <- plot_boxplot_by_param(
      metrics_df, m, param,
      y_label = metric_labels[m],
      x_label = param_labels[param]
    ) + ggplot2::ggtitle(
      paste("Effect of", param_labels[param], "on", metric_labels[m])
    )
    fname <- paste0("02_boxplot_", m, "_by_", param, ".png")
    save_plot(p, file.path(DIR_FIGURES, fname), width = 7, height = 5)
  }
}
# Print one example for the report
plot_boxplot_by_param(metrics_df, "pct_asvs_matching_mock", "filter_mode",
                      y_label = metric_labels["pct_asvs_matching_mock"],
                      x_label = "Filter mode") +
  ggplot2::ggtitle("Effect of filter mode on % ASVs matching mock")
```

### Figure 3: Coefficient plots (model effect sizes)

These show the direction and magnitude of each parameter effect on the link
scale. A coefficient far from zero and whose CI does not include zero indicates
a statistically meaningful effect.

```{r fig-coefficients, fig.height=6, fig.width=9}
for (m in names(models_list)) {
  p <- plot_coefficient(models_list[[m]])
  if (is.null(p)) next
  print(p)
  save_plot(p,
            file.path(DIR_FIGURES, paste0("03_coef_plot_", m, ".png")),
            width = 10, height = 6)
}
```

### Figure 4: Combined overview — all metrics × all parameters (heatmap of marginal means)

```{r fig-heatmap, fig.width=13, fig.height=7}
# Normalise means to [0,1] within each metric for cross-metric comparison
heatmap_df <- mm_df |>
  filter(
  metric %in% names(metric_labels),
  metric != "mean_reads_per_sample"
) |>
  group_by(metric) |>
  mutate(
    norm_mean = (mean - min(mean)) / (max(mean) - min(mean) + 1e-9)
  ) |>
  ungroup() |>
  mutate(
    metric_label = metric_labels[metric],
    param_label  = dplyr::recode(param,
      filter_mode  = "Filter",
      error_model  = "Error model",
      pool_mode    = "Pool",
      omega_raw    = "Omega (ω)",
      band_size    = "Band size"
    )
  )

p_heat <- ggplot2::ggplot(heatmap_df,
       ggplot2::aes(x = param_value, y = metric_label, fill = norm_mean)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(ggplot2::aes(label = round(mean, 2)),
                     size = 3.2, colour = "grey10") +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#D6604D",
    midpoint = 0.5,
    name = "Relative\nmean\n(0–1)"
  ) +
  ggplot2::facet_wrap(~ param_label, scales = "free_x", nrow = 1) +
  ggplot2::labs(
    title    = "Heatmap of parameter effects: normalised marginal means",
    subtitle = "Values shown are raw means; colour is normalised within each metric row",
    x        = "Parameter level",
    y        = NULL
  ) +
  theme_benchmark() +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10),
    axis.text.x = ggplot2::element_text(size = 9)
  )

print(p_heat)
save_plot(p_heat, file.path(DIR_FIGURES, "04_heatmap_overview.png"),
          width = 16, height = 6)
```

---

# PART 2 — Find the best parameter combination

## Composite score: rationale and calculation

### Why two metrics are necessary

Evaluating a DADA2 parameter combination with a single metric is insufficient,
because the two most informative dimensions of mock-community performance
capture fundamentally different aspects of pipeline behaviour:

- **Noise** (contamination): the proportion of observed ASVs that do *not*
  match any sequence in the theoretical mock. These are spurious sequences —
  artefacts of denoising, chimera removal, or sequencing error — that should
  ideally be absent. An **ASV-based** metric is appropriate here because each
  spurious ASV represents a distinct false positive, regardless of how many
  reads support it. A parameter combination that generates many low-abundance
  spurious sequences would go undetected by a read-based metric, but is
  correctly penalised by an ASV-based one.

- **Recovery**: the proportion of total reads in the mock samples that are
  assigned to ASVs which *do* match the theoretical mock. A **read-based**
  metric is appropriate here because it reflects how completely the pipeline
  retains the true signal in terms of sequencing depth. A parameter
  combination that identifies the correct sequences but loses most of their
  reads — due to excessive filtering or chimera removal — would be biologically
  misleading and is correctly penalised by a read-based recovery metric.

Neither metric alone is sufficient. A combination could achieve very low noise
simply by being overly conservative and producing very few ASVs (discarding
real signal in the process), or very high read recovery by being too permissive
and retaining many reads alongside numerous spurious ASVs. Only by considering
both simultaneously can we identify parameter combinations that are
simultaneously clean and complete.

### How the composite score is calculated

For each of the 96 parameter combinations, two raw metrics are extracted from
the mock-only subset of the data:

| Symbol | Metric | Direction |
|---|---|---|
| $A$ | % ASVs not matching theoretical mock | lower is better |
| $B$ | % reads from mock-matching ASVs | higher is better |

Both metrics are **standardised** (z-scored) across all 96 cases to place them
on a common, unit-free scale and remove the effect of their different ranges:

$$z_A = \frac{A - \bar{A}}{\text{SD}(A)}, \qquad z_B = \frac{B - \bar{B}}{\text{SD}(B)}$$

The **composite score** is then computed as:

$$\text{score} = -z_A + z_B$$

This formulation **penalises noise** ($-z_A$: a high noise proportion
increases $z_A$ and subtracts from the score) and **rewards recovery**
($+z_B$: high read recovery increases $z_B$ and adds to the score).
Both components are weighted equally (coefficient = 1), reflecting the
assumption that noise control and signal recovery are of equal biological
importance. This weighting can be adjusted via the `weight_noise` and
`weight_recovery` arguments in `rank_parameter_combinations()`.

Cases are ranked in **descending order** of the composite score (rank 1 =
best). The method is fully transparent and reproducible: every component
is explicit, scale-free, and individually interpretable.

### Tie handling

Because the composite score is continuous, exact ties are possible. Rather
than applying a hard cut-off at rank 5, the selection uses a
**score-threshold** approach: all cases whose composite score is greater than
or equal to the score of the nominally 5th-ranked case are retained. This
avoids arbitrary exclusion of tied combinations.

---

## Compute ranking

```{r compute-ranking}
# Full ranking of all 96 cases.
# See functions file (rank_parameter_combinations) for the complete
# mathematical description of the composite score.
#
# Briefly:
#   z_noise     = z-score of % ASVs not matching mock  (lower = better)
#   z_recovery  = z-score of % reads matching mock     (higher = better)
#   score       = -z_noise + z_recovery
#   rank        = descending rank of score (rank 1 = best)

ranked_df <- rank_parameter_combinations(metrics_df,
                                          weight_noise    = 1,
                                          weight_recovery = 1)

cat("Ranking complete.\n")
cat("Score range: [", round(min(ranked_df$composite_score), 3),
    ",", round(max(ranked_df$composite_score), 3), "]\n\n")

# Top 10 overview
cat("Top 10 parameter combinations:\n")
ranked_df |>
  select(rank, case_id, filter_mode, error_model, pool_mode, omega_raw, band_size,
         pct_asvs_not_matching_mock, pct_reads_matching_mock, composite_score) |>
  head(10) |>
  print()
```

## Top parameter combinations (tie-aware selection)

```{r top-cases}
# assign_rank_groups() labels every case by the 3 highest distinct score values.
# ALL cases sharing a top-3 score receive the same group label (tie-aware).
ranked_df  <- assign_rank_groups(ranked_df)
top_cases  <- ranked_df[ranked_df$rank_group %in% c("1st", "2nd", "3rd"), ]
n_selected <- nrow(top_cases)

cat("\n========== TOP 3 SCORE LEVELS — PARAMETER COMBINATIONS (n =", n_selected, ") ==========\n")
cat("Groups: 1st =", sum(top_cases$rank_group == "1st"),
    "| 2nd =", sum(top_cases$rank_group == "2nd"),
    "| 3rd =", sum(top_cases$rank_group == "3rd"), "\n")

for (i in seq_len(nrow(top_cases))) {
  row <- top_cases[i, ]
  cat(sprintf(
    "\nRank #%d: %s\n  Score: %.4f\n  %% ASVs not matching mock: %.2f%%\n  %% reads matching mock:    %.2f%%\n",
    row$rank,
    make_case_label(row),
    row$composite_score,
    row$pct_asvs_not_matching_mock,
    row$pct_reads_matching_mock
  ))
}

# Compact table for the report
top_cases_table <- top_cases |>
  select(rank, filter_mode, error_model, pool_mode, omega_raw, band_size,
         pct_asvs_not_matching_mock, pct_reads_matching_mock,
         z_noise, z_recovery, composite_score)
print(top_cases_table)
```

## Save ranking tables

```{r save-ranking}
save_csv(ranked_df,  file.path(DIR_TABLES, "05_all_cases_ranked.csv"))
save_csv(top_cases,  file.path(DIR_TABLES, "06_top_parameter_combinations.csv"))
```

---

## Figures — Part 2

### Figure 5: Scatter plot — two decision metrics, top cases highlighted

Each point represents one of the 96 parameter combinations, positioned by its
two decision metrics. The **ideal region is the top-left** corner: low noise
(few spurious ASVs) combined with high recovery (most reads from real mock
sequences). Cases belonging to the top 3 distinct composite score levels are
highlighted; all others are shown in grey.
Text labels are omitted from the plot interior to preserve readability —
parameter details for highlighted cases are shown in the tables and bar plots below.

```{r fig-scatter-ranking, fig.width=11, fig.height=7}
p_scatter <- plot_ranking_scatter(ranked_df,
               n = sum(ranked_df$rank_group %in% c("1st", "2nd", "3rd")))
print(p_scatter)
save_plot(p_scatter, file.path(DIR_FIGURES, "05_ranking_scatter.png"),
          width = 12, height = 7)
```

### Figure 6: Composite score bar plot — top 3 score levels

```{r fig-top-scores, fig.width=10, fig.height=5}
p_scores <- plot_top3_scores(ranked_df)
print(p_scores)
save_plot(p_scores, file.path(DIR_FIGURES, "06_top3_composite_scores.png"),
          width = 11, height = 5)
```

### Figure 7: Dual-metric bar chart — top 3 score levels (raw values)

Shows the actual percentage values for both decision metrics side by side,
making the raw performance of each top combination explicit.

```{r fig-top-dual, fig.width=11, fig.height=6}
p_dual <- plot_top_n_dual_metrics(ranked_df,
            n = sum(ranked_df$rank_group %in% c("1st", "2nd", "3rd")))
print(p_dual)
save_plot(p_dual, file.path(DIR_FIGURES, "07_top3_dual_metrics.png"),
          width = 12, height = 6)
```

### Figure 8: Full ranking — all 96 cases ordered by composite score

```{r fig-full-ranking, fig.width=12, fig.height=9}
# rank_group is already on ranked_df from the top-cases chunk.
# Sort bars: 1st cases at top, then 2nd, then 3rd, then Other (ascending score
# within each group so best bar sits closest to the group boundary).
ranked_df_plot <- ranked_df[order(
  ranked_df$rank_group,           # factor order: 1st < 2nd < 3rd < Other
  ranked_df$composite_score       # ascending within group
), ]

ranked_df_plot$short_label <- paste0(
  ranked_df_plot$filter_mode, "|",
  ranked_df_plot$error_model, "|",
  ranked_df_plot$pool_mode,   "|",
  ranked_df_plot$omega_raw,   "|B",
  ranked_df_plot$band_size
)
# Factor levels control bar order (bottom-to-top in coord_flip)
ranked_df_plot$short_label <- factor(ranked_df_plot$short_label,
                                      levels = ranked_df_plot$short_label)

rank_colors <- c(
  "1st"   = "#C73E1D",
  "2nd"   = "#F4831F",
  "3rd"   = "#F7C948",
  "Other" = "#B0C4DE"
)
# Pull the stored score values from the attribute set by assign_rank_groups()
s <- attr(ranked_df, "top3_scores")
if (is.null(s)) s <- sort(unique(ranked_df$composite_score), decreasing = TRUE)[1:3]
rank_labels <- c(
  "1st"   = sprintf("1st (score = %.3f)", s[1]),
  "2nd"   = sprintf("2nd (score = %.3f)", s[2]),
  "3rd"   = sprintf("3rd (score = %.3f)", s[3]),
  "Other" = "Other"
)

p_full <- ggplot2::ggplot(ranked_df_plot,
       ggplot2::aes(x = composite_score, y = short_label, fill = rank_group)) +
  ggplot2::geom_col(width = 0.7, alpha = 0.85) +
  ggplot2::scale_fill_manual(
    values = rank_colors,
    labels = rank_labels,
    name   = NULL
  ) +
  ggplot2::labs(
    title    = "Composite score ranking — all 96 parameter combinations",
    subtitle = "Score = \u2013z(noise) + z(recovery); higher = better | Top 3 score levels highlighted",
    x        = "Composite score",
    y        = "Parameter combination"
  ) +
  theme_benchmark() +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 7),
    axis.text.x = ggplot2::element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )

print(p_full)
save_plot(p_full, file.path(DIR_FIGURES, "08_full_ranking_all_cases.png"),
          width = 12, height = 16)
```

---

# Session information

```{r session-info}
sessionInfo()
```

---

# Appendix: Metric definitions

| Metric | Definition | Computed from |
|---|---|---|
| `pct_chimeric_reads` | Reads assigned to chimeric ASVs / total reads in mock samples × 100 | `$bim` + `$st` (pre-removal table) |
| `mean_reads_per_sample` | Mean of per-sample read totals across mock samples | `$st_nochim` |
| `total_asvs` | Number of distinct ASVs with ≥ 1 read in any mock sample | `$st_nochim` |
| `pct_asvs_matching_mock` | % of observed ASVs (in mock samples) that exactly match a theoretical mock sequence (IUPAC-aware) | `$st_nochim` + FASTA |
| `pct_asvs_not_matching_mock` | 100 − `pct_asvs_matching_mock` (noise metric, Part 2) | derived |
| `pct_reads_matching_mock` | % of total reads in mock samples coming from mock-matching ASVs (recovery metric, Part 2) | `$st_nochim` + FASTA |

**IUPAC matching:** an ASV sequence matches a reference sequence if (a) they
are the same length, and (b) at every position, the observed nucleotide belongs
to the set represented by the reference IUPAC code. This correctly handles
the single `V` (= A, C, or G) ambiguity in the ESBL15 reference sequence.

**Ranking score:** composite_score = −z(noise) + z(recovery), where z-scores
are computed across all 96 cases. A weight of 1 is applied to each component
(symmetric weighting). To change the weighting, modify the `weight_noise` and
`weight_recovery` arguments in `rank_parameter_combinations()`.
