---
title: "02 — Community Analyses"
subtitle: "DADA2 Pipeline · Parameter Sensitivity & Sample Type Assessment"
author: "Andrés Catalán Tatay"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 14
    fig_height: 8
    df_print: paged
---

<!--
================================================================================
  Project : fliC Amplicon Sequencing for High-Resolution Profiling of E. coli
            in the Infant Gut Microbiome
  Author  : Andrés Catalán Tatay
  Part    : Analysis — Step 02
  Purpose : Community-level analysis of all 96 DADA2 pipeline parameter
            combinations. Covers ASV sharing, distance matrices, PERMANOVA,
            PCoA ordination, and dendrogram heatmaps.
  Inputs  : data/cases/          — 96 RDS files (one per parameter combination)
  Outputs : results/02_community_analyses/
              01_asv_sharing/    — ASV sharing bar charts and CSV tables
              02_pcoa/           — PCoA plots (Jaccard + Aitchison, 6 panels)
              03_dendrogram/     — Hierarchical clustering heatmaps
              04_permanova/      — PERMANOVA results plot and CSV
              05_combined/       — Summary figure combining all panels
              distances/         — Cached distance matrix RDS files
  Companion: 02_community_analyses_functions.R
================================================================================
-->

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  dpi       = 150,
  fig.align = "center",
  out.width = "100%"
)
```

---

# Overview

This notebook performs a complete community-level analysis on **all 96 DADA2
pipeline cases**, covering every combination of five parameters:

| Parameter         | Levels |
|-------------------|--------|
| **Filtering**     | Loose · Strict |
| **Pooling**       | FALSE · pseudo · TRUE |
| **Omega (χ²)**   | O1e-40 · O1e-70 · O1e-100 · O1e-200 |
| **Band size**     | 32 · 64 |
| **Error model**   | Binned · PacBio |
| **Sample type**   | Individual · Positive (mockmix) · Negative |

2 × 3 × 4 × 2 × 2 = **96 cases**


# 0 · Packages & Setup

```{r install-check, results='hide'}
# ── Install any missing packages ───────────────────────────────────────────────
bioc_pkgs <- c("phyloseq", "ComplexHeatmap")
cran_pkgs <- c("ggplot2", "dplyr", "tidyr", "tibble", "vegan", "ape",
               "patchwork", "RColorBrewer", "stringr", "ggrepel",
               "scales", "circlize", "grid", "parallel")

all_pkgs <- c(bioc_pkgs, cran_pkgs)
miss     <- all_pkgs[!vapply(all_pkgs, requireNamespace,
                              logical(1L), quietly = TRUE)]
if (length(miss) > 0L) {
  miss_bioc <- intersect(miss, bioc_pkgs)
  miss_cran <- setdiff(miss, bioc_pkgs)
  if (length(miss_bioc) > 0L) {
    if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager")
    BiocManager::install(miss_bioc, ask = FALSE, update = FALSE)
  }
  if (length(miss_cran) > 0L)
    install.packages(miss_cran, dependencies = TRUE)
}
invisible(lapply(all_pkgs, library, character.only = TRUE))
```

```{r source-functions}
# ── Load the companion functions file ─────────────────────────────────────────
fn_file <- "02_community_analyses_functions.R"
if (!file.exists(fn_file))
  stop("Functions file not found: '", fn_file,
       "'\nPlace it in the same folder as this Rmd.")
source(fn_file)
message("Functions loaded from: ", fn_file)
```

```{r define-paths}
# ── Paths ─────────────────────────────────────────────────────────────────────
# All paths are relative to the project root.
# Set your working directory to the project root before knitting.
INPUT_DIR  <- "data/cases"
OUTPUT_DIR <- "results/02_community_analyses"

# ── Parallelism ───────────────────────────────────────────────────────────────
# Windows does not support mclapply → force sequential
N_CORES <- if (.Platform$OS.type == "windows") 1L else
             max(1L, parallel::detectCores() - 1L)
message("Parallel cores available: ", N_CORES)

# ── Create output sub-folders ─────────────────────────────────────────────────
subdirs <- c("01_asv_sharing", "02_pcoa",
             "03_dendrogram",  "04_permanova", "05_combined", "distances")
invisible(lapply(file.path(OUTPUT_DIR, subdirs),
                 dir.create, recursive = TRUE, showWarnings = FALSE))

cat("Input : ", INPUT_DIR,  "\n")
cat("Output: ", OUTPUT_DIR, "\n")
```

---

# 1 · Load Cases (Change A – all 96)

> **Change A** – All 96 cases are loaded without any filename filtering.
> Expected result: **96 cases** (2 filter × 3 pool × 4 omega × 2 band × 2 error model).

```{r load-cases}
cases <- load_all_cases(INPUT_DIR, n_cores = N_CORES)
cat("\nCases loaded:", length(cases), " (expected: 96)\n\n")

# Show the full parameter grid
param_df <- do.call(rbind,
                    lapply(cases, function(x) as.data.frame(x$params,
                                                             stringsAsFactors = FALSE)))
cat("Parameter grid (cases loaded):\n")
print(with(param_df,
           table(Filter_Mode = Filter_Mode, Pool = Pool,
                 Omega = Omega, Band_Size = Band_Size, Error_Model = Error_Model)))
```

---

# 2 · Remove Pure Mocks & Enrich Metadata (Change B)

> Pure mock controls (`S_KPC*`, `S_OXA*`, `S_E15`, `S_E25`)
> are removed from every phyloseq object **before** classification.
> Only MOCKMIX community standards remain as **Positive**.
> Real patient/environment samples are labelled **Individual**.
>
> **IMPORTANT:** After running this chunk, inspect the printed sample lists
> and verify that no unwanted names appear in any category.

```{r enrich-remove-mocks}
# Enrich metadata (adds Filter_Mode, Pool, Omega, Band_Size, Error_Model,
# sample_type) AND removes pure mock controls from each phyloseq.
ps_list <- enrich_and_clean_cases(cases, n_cores = N_CORES)

cat("\n── Sample counts per case after pure mock removal ──\n")
n_per_case <- vapply(ps_list, nsamples, integer(1L))
cat("  Min:", min(n_per_case),
    " Max:", max(n_per_case),
    " Mean:", round(mean(n_per_case), 1), "\n\n")
```

---

# 3 · Rarefaction (Change C – fixed at 1 500 reads)

> **Change C** – Sequencing depth is normalised using `rarefy_even_depth()`
> to a **fixed depth of 1 500 reads** per sample.
>
> **Handling low-depth samples (any type):**
> - Samples with reads ≥ 1 500 are rarefied normally.
> - Samples with reads < 1 500 are **kept unrarefied** (they would
>   otherwise be dropped entirely, losing that sample from the analysis).
>   A warning is printed for any such sample.
>
> A fixed seed is used for full reproducibility.

```{r rarefy}
# Fixed rarefaction depth of 1 500 reads (set manually, not derived from data)
min_depth <- 1500L
cat("Rarefaction depth: ", min_depth, " reads (fixed by user)\n\n")

# Rarefy all 96 cases (seed = 42 for reproducibility)
ps_list_rarefied <- rarefy_cases(ps_list,
                                  min_depth = min_depth,
                                  seed      = 42L)
cat("\nRarefaction complete. Sample counts after rarefaction:\n")
n_after <- vapply(ps_list_rarefied, nsamples, integer(1L))
cat("  Min:", min(n_after),
    " Max:", max(n_after),
    " Mean:", round(mean(n_after), 1), "\n")
```

---

# 4 · Merge All 96 Cases

```{r merge-cases}
# Merge all 96 rarefied cases; sample names are already unique (__caseN suffix)
ps_all   <- merge_all_cases(ps_list_rarefied)
meta_all <- as.data.frame(sample_data(ps_all))

cat("\n=== Merged phyloseq ===\n")
print(ps_all)

# Verify sample classification
cat("\n── Sample type distribution (after pure mock removal & rarefaction) ──\n")
cat("  Negative   = S_TRIS (buffer-only controls)\n")
cat("  Positive   = MOCKMIX community standards\n")
cat("  Individual = S_<number> (real patient/environment samples)\n\n")
print(table(Sample_type = meta_all$sample_type))

# Print representative names per type for manual verification
for (st in c("Individual", "Positive", "Negative")) {
  ex <- unique(sub("__case\\d+$", "",
                   rownames(meta_all)[meta_all$sample_type == st]))
  cat("\n  ", st, "– examples: ", paste(head(ex, 8L), collapse = ", "), "\n")
}

# Parameter distributions
cat("\nParameter distributions in merged phyloseq:\n")
cat("  Pool        :\n"); print(table(meta_all$Pool))
cat("  Filter_Mode :\n"); print(table(meta_all$Filter_Mode))
cat("  Omega       :\n"); print(table(meta_all$Omega))
cat("  Band_Size   :\n"); print(table(meta_all$Band_Size))
cat("  Error_Model :\n"); print(table(meta_all$Error_Model))
```

---

# 5 · ASV Origin Analysis

> All 96 cases contribute to every comparison. No parameter is held
> fixed or excluded during ASV sharing analysis.
>
> Now includes comparisons for all 5 DADA2 parameters:
> **Filter_Mode / Pool / Omega / Band_Size / Error_Model**

```{r asv-sharing-compute}
# All 5 parameters (Change D)
params_sharing <- c("Filter_Mode", "Pool", "Omega", "Band_Size", "Error_Model")

sharing_results <- setNames(
  lapply(params_sharing, function(p) {
    summarise_asv_sharing(ps_all, p)
  }),
  params_sharing
)

# Print counts per origin category
for (p in params_sharing) {
  cat("\n── ASV sharing by '", p, "' ──\n", sep = "")
  cnt <- sort(table(sharing_results[[p]]$origin), decreasing = TRUE)
  print(cnt)
}
```

```{r asv-sharing-plot, fig.width=22, fig.height=12}
# Two rows: Filter_Mode / Pool / Omega (top) + Band_Size / Error_Model (bottom)
p_sharing <- plot_asv_sharing_all(sharing_results)
print(p_sharing)

save_plot(p_sharing,
          file.path(OUTPUT_DIR, "01_asv_sharing",
                    "ASV_sharing_all_params.png"),
          width = 22, height = 12)
```

```{r asv-sharing-tables, results='hide'}
for (p in params_sharing) {
  write.csv(sharing_results[[p]],
            file.path(OUTPUT_DIR, "01_asv_sharing",
                      paste0("ASV_sharing_", p, ".csv")),
            row.names = FALSE)
}
message("ASV sharing tables saved.")
```

---

# 6 · Distance Matrices

> Jaccard and Aitchison distances computed on the rarefied, merged
> phyloseq (all 96 cases, pure mocks removed).

```{r compute-distances}
# Distance matrices with disk caching
# Set FORCE_RECOMPUTE <- TRUE to ignore cache and recalculate
DIST_CACHE_DIR  <- file.path(OUTPUT_DIR, "distances")
FORCE_RECOMPUTE <- FALSE

dist_list   <- compute_distances_parallel(
  ps              = ps_all,
  cache_dir       = DIST_CACHE_DIR,
  n_cores         = N_CORES,
  force_recompute = FORCE_RECOMPUTE
)
d_jaccard   <- dist_list$Jaccard
d_aitchison <- dist_list$Aitchison
```

---

# 7 · PERMANOVA (Change E – all 6 parameters)

> **Correct experimental design:** For each of the 6 parameters, all
> other parameters are fixed at their reference level so each biological
> sample appears exactly **once** per level of the target parameter.
> This gives ~100 samples per test — statistically clean and fast.
>
> Reference levels:
> - Pool = pseudo · Filter_Mode = Loose · Omega = O1e-200
> - Band_Size = 32 · Error_Model = Binned

```{r permanova-source-check}
source(fn_file)
message("Functions re-loaded: ", fn_file)
```

```{r permanova-run}
perm_tidy <- run_permanova_by_param(
  dist_list    = dist_list,
  ps           = ps_all,
  permutations = 999L,
  ref_pool     = "pseudo",
  ref_filter   = "Loose",
  ref_omega    = "O1e-200",
  ref_band     = "32",
  ref_err      = "Binned"
)

# Split for downstream PCoA annotation
perm_jac_tidy <- perm_tidy |> dplyr::filter(dist_method == "Jaccard")
perm_ait_tidy <- perm_tidy |> dplyr::filter(dist_method == "Aitchison")

cat("\n=== PERMANOVA results (all 6 parameters) ===\n")
print(perm_tidy[, c("dist_method","tested_param","Term",
                     "R2","F_stat","p_value","Significance")])
```

```{r permanova-plot, fig.width=10, fig.height=7}
p_perm <- plot_permanova_results(perm_tidy)
print(p_perm)

save_plot(p_perm,
          file.path(OUTPUT_DIR, "04_permanova", "PERMANOVA_results.png"),
          width = 10, height = 7)

write.csv(perm_tidy,
          file.path(OUTPUT_DIR, "04_permanova", "PERMANOVA_all.csv"),
          row.names = FALSE)
```

---

# 8 · PCoA Ordination (Change D – 6 panels)

> Two rows of three panels each:
> Row 1: sample_type / Pool / Filter_Mode
> Row 2: Omega / Band_Size / Error_Model

## 8a · Jaccard

```{r pcoa-jaccard-compute}
message("Running PCoA – Jaccard ...")
pcoa_jac <- run_pcoa(d_jaccard, ps_all, dist_label = "Jaccard")
cat("Variance explained (PC1-4):",
    paste0(attr(pcoa_jac, "pct"), "%"), "\n")
```

```{r pcoa-jaccard-6panel, fig.width=24, fig.height=12}
p_pcoa_jac <- plot_pcoa_multi(
  pcoa_jac, perm_df = perm_jac_tidy,
  params = c("sample_type","Pool","Filter_Mode","Omega","Band_Size","Error_Model")
)
print(p_pcoa_jac)
save_plot(p_pcoa_jac,
          file.path(OUTPUT_DIR, "02_pcoa", "PCoA_Jaccard_6panels.png"),
          width = 24, height = 12)
```

## 8b · Aitchison

```{r pcoa-aitchison-compute}
message("Running PCoA – Aitchison ...")
pcoa_ait <- run_pcoa(d_aitchison, ps_all, dist_label = "Aitchison")
cat("Variance explained (PC1-4):",
    paste0(attr(pcoa_ait, "pct"), "%"), "\n")
```

```{r pcoa-aitchison-6panel, fig.width=24, fig.height=12}
p_pcoa_ait <- plot_pcoa_multi(
  pcoa_ait, perm_df = perm_ait_tidy,
  params = c("sample_type","Pool","Filter_Mode","Omega","Band_Size","Error_Model")
)
print(p_pcoa_ait)
save_plot(p_pcoa_ait,
          file.path(OUTPUT_DIR, "02_pcoa", "PCoA_Aitchison_6panels.png"),
          width = 24, height = 12)
```

---

# 9 · Dendrogram Heatmaps (Change D – 6 annotation strips)

> The dendrogram shows sample clustering by distance.
> Annotation strips on the right now show all 6 variables:
> sample_type · Pool · Filter · Omega · Band size · Error model.
> PERMANOVA R² and significance are embedded in the figure title.

## 9a · Jaccard

```{r dendro-jaccard, fig.width=18, fig.height=13}
ht_jac <- plot_dendrogram_heatmap(
  dist_obj       = d_jaccard,
  ps             = ps_all,
  dist_label     = "Jaccard",
  perm_df        = perm_jac_tidy,
  cluster_method = "ward.D2"
)
ComplexHeatmap::draw(ht_jac,
                     merge_legend           = TRUE,
                     heatmap_legend_side    = "right",
                     annotation_legend_side = "right")
cap_jac <- attr(ht_jac, "perm_caption")
if (!is.null(cap_jac) && nchar(cap_jac) > 0L)
  cat("\n**PERMANOVA:**", cap_jac, "\n")
save_heatmap(ht_jac,
             file.path(OUTPUT_DIR, "03_dendrogram", "Dendrogram_Jaccard.png"),
             width = 18, height = 13)
```

## 9b · Aitchison

```{r dendro-aitchison, fig.width=18, fig.height=13}
ht_ait <- plot_dendrogram_heatmap(
  dist_obj       = d_aitchison,
  ps             = ps_all,
  dist_label     = "Aitchison",
  perm_df        = perm_ait_tidy,
  cluster_method = "ward.D2"
)
ComplexHeatmap::draw(ht_ait,
                     merge_legend           = TRUE,
                     heatmap_legend_side    = "right",
                     annotation_legend_side = "right")
cap_ait <- attr(ht_ait, "perm_caption")
if (!is.null(cap_ait) && nchar(cap_ait) > 0L)
  cat("\n**PERMANOVA:**", cap_ait, "\n")
save_heatmap(ht_ait,
             file.path(OUTPUT_DIR, "03_dendrogram", "Dendrogram_Aitchison.png"),
             width = 18, height = 13)
```

---

# 10 · Combined Summary Figure

> Row 1: Jaccard PCoA (6 panels, 2 rows)
> Row 2: Aitchison PCoA (6 panels, 2 rows)
> Row 3: PERMANOVA (lollipop + p-value table) | ASV sharing

```{r combined, fig.width=24, fig.height=26}
p_combined <- (p_pcoa_jac / p_pcoa_ait / (p_perm | p_sharing)) +
  plot_annotation(
    title    = "Community Analysis Summary – All 96 DADA2 Cases",
    subtitle = paste0(
      nsamples(ps_all), " samples  |  ",
      ntaxa(ps_all),    " ASVs  |  ",
      length(cases),    " pipeline cases",
      "  |  rarefied to 1 500 reads (fixed)"
    ),
    tag_levels = "A",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40"),
      plot.tag      = element_text(face = "bold", size = 12)
    )
  )
print(p_combined)
save_plot(p_combined,
          file.path(OUTPUT_DIR, "05_combined", "Community_summary.png"),
          width = 24, height = 26)
```

---

# 11 · Session Info

```{r session-info}
sessionInfo()
```
