---
title: "05 — Mock Distributions (Check Mock Analysis)"
author: "Andrés Catalán Tatay"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

<!--
================================================================================
  Project : fliC Amplicon Sequencing for High-Resolution Profiling of E. coli
            in the Infant Gut Microbiome
  Author  : Andrés Catalán Tatay
  Part    : Analysis — Step 05
  Purpose : Quality control of mock community standards.
            Assesses how well the experimental mock samples (Mock Mix and
            Singular Mocks) match the theoretical mock composition for both
            the BEST and STANDARD pipelines.
            Uses Bray-Curtis similarity against the theoretical reference
            instead of checkZymoBiomics() (see Overview for rationale).
  Inputs  : data/cases/case_049__F-Strict__Err-Binned__Pool-FALSE__O1e-200__B32.rds
             data/cases/case_031__F-Loose__Err-PacBio__Pool-FALSE__O1e-40__B32.rds
             data/mock/ps_mock_trimmed_4.rds  — theoretical mock reference
  Outputs : results/05_mock_qc/
              05_Mock_CheckZymo_case049_BEST.png/.pdf
              05_Mock_CheckZymo_case031_STANDARD.png/.pdf
  Companion: 05_Mock_Distributions_Functions.R
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = TRUE,
  fig.width  = 12,
  fig.height = 7
)
```

---

## Overview

**"Check Mock"** quality-control analysis — **full read counts (no rarefaction)**.

Both the BEST and STANDARD pipelines are processed with identical code.
Mock Mix samples (`S_Mock_mix*`) and Singular Mock samples (`S_KPC1`, `S_Oxa`,
`S_E15`, `S_E25`) are analysed **together** in a single combined object and appear
in the same comparison plots against the theoretical mock.

The central question:

> *Is the composition in experimental mock standards similar to the
> theoretically expected composition?*



---

## 0 · Source Helper Functions

```{r source-functions}
# ── Load the companion functions file ─────────────────────────────────────────
fn_file <- "05_Mock_Distributions_Functions.R"
if (!file.exists(fn_file))
  stop("Functions file not found: '", fn_file,
       "'\nPlace it in the same folder as this Rmd.")
source(fn_file)
```

---

## 1 · File Paths & Parameters

```{r file-paths}
# ── Paths ─────────────────────────────────────────────────────────────────────
# All paths are relative to the project root.
# Set your working directory to the project root before knitting.

# Shared paths
cases_dir  <- "data/cases"
theo_path  <- "data/mock/ps_mock_trimmed_4.rds"
fig_dir    <- "results/05_mock_qc"

# Case-specific paths
case_path_BEST     <- file.path(cases_dir,
  "case_049__F-Strict__Err-Binned__Pool-FALSE__O1e-200__B32.rds")

case_path_STANDARD <- file.path(cases_dir,
  "case_031__F-Loose__Err-PacBio__Pool-FALSE__O1e-40__B32.rds")

# Shared analysis parameter (applied identically to both pipelines)
MIN_SIM <- 99.9  # minimum % identity to assign an ASV to a theoretical taxon
```

---

## 2 · Load Shared Theoretical Reference

```{r load-theoretical}
ps_theoretical <- load_theoretical_mock(theo_path)
ps_theoretical
```

```{r diag-theoretical}
diagnose_ps(ps_theoretical, "Theoretical mock reference")
```

```{r show-theoretical-taxa}
cat("Theoretical taxa names:\n")
print(taxa_names(ps_theoretical))

cat("\nTheoretical tax_table:\n")
if (!is.null(access(ps_theoretical, "tax_table", errorIfNULL = FALSE))) {
  print(as(tax_table(ps_theoretical), "matrix"))
} else {
  cat("  (no tax_table)\n")
}
```

---

# CASE: BEST  (case_049__F-Strict__Err-Binned__Pool-FALSE__O1e-200__B32)

## 3 · Load Case — BEST

```{r load-case-best}
ps_case_BEST <- load_case_phyloseq(case_path_BEST)
ps_case_BEST
```

```{r diag-case-best}
diagnose_ps(ps_case_BEST, "BEST -- Full case phyloseq (before extraction)")
```

---

## 4 · Inspect Sample Metadata — BEST

```{r inspect-metadata-best}
cat("All sample names in BEST case object:\n")
print(sample_names(ps_case_BEST))
```

---

## 5 · Extract All Mock Samples (Mock Mix + Singular) — BEST

```{r extract-all-mocks-best}
ps_all_mocks_BEST <- extract_all_mock_samples(ps_case_BEST)
```

```{r diag-all-mocks-best}
diagnose_ps(ps_all_mocks_BEST, "BEST -- After combined mock sample extraction")
```

```{r show-all-mock-counts-best}
cat("BEST -- All mock sample read counts:\n")
print(sample_sums(ps_all_mocks_BEST))
```

---

## 6 · Run Check Mock Analysis — BEST

```{r run-check-mock-best, cache=TRUE}
output_dat_BEST <- run_custom_check_mock(
  ps_mocks       = ps_all_mocks_BEST,
  ps_theoretical = ps_theoretical,
  min_sim        = MIN_SIM
)
```

---

# CASE: STANDARD  (case_031__F-Loose__Err-PacBio__Pool-FALSE__O1e-40__B32)

## 7 · Load Case — STANDARD

```{r load-case-standard}
ps_case_STANDARD <- load_case_phyloseq(case_path_STANDARD)
ps_case_STANDARD
```

```{r diag-case-standard}
diagnose_ps(ps_case_STANDARD, "STANDARD -- Full case phyloseq (before extraction)")
```

---

## 8 · Inspect Sample Metadata — STANDARD

```{r inspect-metadata-standard}
cat("All sample names in STANDARD case object:\n")
print(sample_names(ps_case_STANDARD))
```

---

## 9 · Extract All Mock Samples (Mock Mix + Singular) — STANDARD

```{r extract-all-mocks-standard}
ps_all_mocks_STANDARD <- extract_all_mock_samples(ps_case_STANDARD)
```

```{r diag-all-mocks-standard}
diagnose_ps(ps_all_mocks_STANDARD, "STANDARD -- After combined mock sample extraction")
```

```{r show-all-mock-counts-standard}
cat("STANDARD -- All mock sample read counts:\n")
print(sample_sums(ps_all_mocks_STANDARD))
```

---

## 10 · Run Check Mock Analysis — STANDARD

```{r run-check-mock-standard, cache=TRUE}
output_dat_STANDARD <- run_custom_check_mock(
  ps_mocks       = ps_all_mocks_STANDARD,
  ps_theoretical = ps_theoretical,
  min_sim        = MIN_SIM
)
```

---

## 11 · Build Shared Colour Palette

A single palette is built from the union of all taxa across both cases so the
same ASV always gets the same colour in every plot.

```{r build-shared-palette}
shared_pal <- build_shared_palette(
  output_dat_BEST,
  output_dat_STANDARD
)
```

---

# RESULTS — BEST

## 12 · Correlation Table — BEST

```{r correlation-table-best}
cortable_BEST <- get_correlation_table(output_dat_BEST)
cortable_BEST
```

> **Interpretation:** Values close to **1.0** indicate the experimental mock
> closely matches the theoretical composition.

---

## 13 · Species-Level Merged Data — BEST

```{r species-level-best}
ps_species_BEST <- output_dat_BEST$ps_species
ps_species_BEST

cat("\nSpecies / taxa in merged object (BEST):\n")
print(taxa_names(ps_species_BEST))
```

---

## 14 · Plot Composition — BEST

```{r plot-composition-best, fig.width=12, fig.height=7}
p_BEST <- plot_check_mock(output_dat_BEST, shared_pal = shared_pal,
                          title = "BEST Pipeline -- Mock Composition vs Theoretical")
p_BEST
```

---

## 15 · Save Figure — BEST

```{r save-figures-best}
save_figure(
  plot_obj  = p_BEST,
  base_name = "05_Mock_CheckZymo_case049_BEST",
  out_dir   = fig_dir,
  width     = 12,
  height    = 7
)
```

---

# RESULTS — STANDARD

## 16 · Correlation Table — STANDARD

```{r correlation-table-standard}
cortable_STANDARD <- get_correlation_table(output_dat_STANDARD)
cortable_STANDARD
```

> **Interpretation:** Values close to **1.0** indicate the experimental mock
> closely matches the theoretical composition.

---

## 17 · Species-Level Merged Data — STANDARD

```{r species-level-standard}
ps_species_STANDARD <- output_dat_STANDARD$ps_species
ps_species_STANDARD

cat("\nSpecies / taxa in merged object (STANDARD):\n")
print(taxa_names(ps_species_STANDARD))
```

---

## 18 · Plot Composition — STANDARD

```{r plot-composition-standard, fig.width=12, fig.height=7}
p_STANDARD <- plot_check_mock(output_dat_STANDARD, shared_pal = shared_pal,
                              title = "STANDARD Pipeline -- Mock Composition vs Theoretical")
p_STANDARD
```

---

## 19 · Save Figure — STANDARD

```{r save-figures-standard}
save_figure(
  plot_obj  = p_STANDARD,
  base_name = "05_Mock_CheckZymo_case031_STANDARD",
  out_dir   = fig_dir,
  width     = 12,
  height    = 7
)
```

---

## 20 · Session Info

```{r session-info}
sessionInfo()
```
